Von Neumann-Morgenstern and Savage Theorems for Causal Decision Making

### 2020-09-10 Artificial_Intelligence_Journal/Draft/AIJ_Draft.tex
draft para AIJ

\RequirePackage{fix-cm}

\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
%
%
\usepackage{mathptmx}      
\usepackage[utf8]{inputenc}
\usepackage[pdftex]{color,graphicx}
\usepackage{listings}
\usepackage{breakcites}
\usepackage{dsfont}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{amssymb,amsmath,latexsym}
\usepackage[margin=1.5cm]{geometry}
\usepackage{enumitem}
%\bibliographystyle{stylename}
%\usepackage{fancyhdr}
%\pagestyle{fancy}
\usepackage{tikz}
\usetikzlibrary{trees}
\usetikzlibrary{calc}
\usepackage{color}
\usepackage{todonotes}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\begin{document}

@@

%\titlerunning{Short form of title}        % if too long for running head
%
%\author{Mauricio Gonzalez-Soto         \and
%        Luis E. Sucar \and
%        Hugo J. Escalante
%}
\author{Anonymous}
%\authorrunning{Short form of author list} % if too long for running head

%\institute{Mauricio Gonzalez-Soto \at
%              Coordinaci\'on de Ciencias Computacionales, Instituto Nacional de Astrof\'isica \'Optica y Electr\'onica \\
%              Luis Enrique Erro 1, Santa Maria Tonanzintla, Puebla\\
%              Mexico\\
%              \email{mauricio@inaoep.mx}           \\
%              \url{https://orcid.org/0000-0003-2668-9013}        
%           \and
%           Luis E. Sucar \at
%              Coordinaci\'on de Ciencias Computacionales, Instituto Nacional de Astrof\'isica \'Optica y Electr\'onica \\
%              Luis Enrique Erro 1, Santa Maria Tonanzintla, Puebla\\
%              M\'exico\\
%              \email{esucar@inaoep.mx}\\
%              \url{https://orcid.org/0000-0002-3685-5567}
%           \and
%           Hugo J. Escalante \at
%           Coordinaci\'on de Ciencias Computacionales, Instituto Nacional de Astrof\'isica \'Optica y Electr\'onica \\
%              Luis Enrique Erro 1, Santa Maria Tonanzintla, Puebla\\
%              M\'exico\\
%              \email{hugojair@inaoep.mx}\\
%              \url{https://orcid.org/0000-0003-4603-3513}
%}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle
% modificar abstract para reflejar cambios en el paper
\begin{abstract} 
Causal reasoning is a fundamental aspect of intelligent reasoning since it allows to consider interventions and counterfactuals. Decision making under uncertainty has been well studied when information is considered at the associative (probabilistic) level. The classical Theorems of von Neumann-Morgenstern and Savage provide a formal criterion for rational choice using purely associative information, which form the basis of many learning algorithms such as those used in Reinforcement Learning. In this work, we consider decision problems in which available actions and consequences are causally connected. We define a Causal Decision Problem and within this framework we state a previous result from J. Pearl, which relies on a known causal model and thus showing that it can be considered as a causal version of the classical von Neumann-Morgenstern Theorem. Furthermore, we consider the case when the causal mechanism that controls the environment is unknown to the decision maker, and propose and prove a causal version of Savage's Theorem. Then, we describe two applications for which these theorems provide theoretical foundations: causal games and optimal action learning in causal environments. These results highlight the importance of causal models in decision making and the variety of potential applications.

\keywords{Causality \and Decision Making \and Game Theory}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}
\tableofcontents
\newpage
\section{Introduction}
\label{intro}
Causal reasoning is a constant element in our lives as it is in human nature to constantly ask \textit{why} \citep{spirtes2000causation,pearl2018why}. Looking for causes is an everyday task and, in fact, causal reasoning is to be found at the very core of our minds \citep{tversky1977causal,waldmann2013causal,danks2014unifying,neil2019causality}. It has been argued that the brain is a causal inference machine which uses \textit{effects} to figure out \textit{causes} in order to engage with the world \citep{friston2010free,clark2015surfing}. 
Furthermore, acting in the world is conceived by human beings as \textit{intervening} the world and in fact humans are able to learn and use causal relations while making choices \citep{ tversky1980causal, lagnado2007beyond, hagmayer2008causal, hagmayer2009decision, hagmayer2017causality}.

@@

%For example, as told by \cite{bornstein2016artificial}, at the University of Pittsburgh Medical Center a team of researchers tried to use Machine Learning to predict whether pneumonia patients might develop severe complications. For this purpose they trained Neural Networks and Decision Trees using the hospital's own data. Neural Networks outperformed Decision Trees, but only by studying the decisions made by the latter the doctors did find out that the algorithms instructed doctors to send home pneumonia patients who already had asthma, despite the fact that asthma sufferers are known to be extremely vulnerable to complications. The problem relied in the training data, because the hospital policy was to automatically send asthma sufferers with pneumonia to intensive care, and this policy worked so well that asthma sufferers almost never developed severe complications. It was only through the interpretability of the Decision Trees that the doctors didn't send asthma patients with pneumonia home to a certain death.

%Methods based in Deep Neural Networks aren't supposed to explain why a certain output was produced since those methods are based in \textit{parallel distributed representations} and the same goes for any other learning algorithm that uses Deep Neural Networks, like Deep Reinforcement learning; when AlphaGo \citep{silver2017mastering} defeated the world-champion, humanity didn't learn anything new about the game of Go, because the algorithm was not designed to explain its moves, although it was a very sophisticated method for achieving it. Using only associations between variables it is not possible to infer which one is the cause and which one the effect, something extra is required. The fundamental difference between causal and associative models is that causal models consider \textit{doing} over \textit{observing} \citep{pearl2018why}. Once a causal relation has been established between two events, performing, or intervening, over a cause allows one to predict the outcome in a more robust way than if using only correlations between observations.

@@

%The impact of Causal Inference goes beyond Machine Learning and Computer Science. For example, economists are interested in understanding what did certain public policy caused \citep{athey2017impact} or how human decision makers that show \textit{inconsistent} preferences over time could be oriented if the causal consequences of this inconsistent behavior is introduced into a decision-making model \citep{albers2016motivating}. In complex adaptive systems, causal relations can be used to clarify the complex interactions between agents in the system \citep{abbott2017complex} as well as for prediction and planning \citep{hunt2016ants,brock2018causality}.

%cerrar esta sección sobre cómo mi trabajo impacta en la motivación

@@

\subsection{Organization of the paper}
\label{organization}
This paper is organized as follows: first, in Section \ref{classical_theorems} we present the classical decision making theorems which rely on associative information in their formulation: Theorems \ref{vNM} and \ref{savage}; then, in Section \ref{causal_decision_problems} we present a brief introduction to Causation and Causal Graphical Models. In Section \ref{related_work} we present previous approaches both to Causal Decision Making as well as its applications to interactive optimal action learning: causal reinforcement learning. In Section \ref{causal_results} we state a von Neumann-Morgenstern Theorem for Causal Decision Making that is due to Pearl, which assumes a known causal structure. Our main result is Theorem \ref{causal_savage} which extends Pearl's result to the case of an unknown causal structure. Finally two applications of Theorem \ref{causal_savage} are described. First, a causal analogue of Nash Equilibrium for games in which actions and consequences are causally connected is defined. Then, in Section \ref{optimal_action} we describe some optimal-action learning methods for causal environments which are able to learn an \textit{optimal intervention} under different assumptions and conditions; in particular, we describe and provide an extension for one of such works. Finally, in Section \ref{limitations} we comment on the limitations of the assumptions we consider. Finally, in Section \ref{conclusion} we summarize the main contributions of our work.

@@

\section*{Declarations}

\section*{Funding}
The present research has been funded by the Consejo Nacional de Ciencia y Tecnolog\'ia (CONACyT).

\section*{Availability of data and material}
Not applicable

\section*{Code availability}
Not applicable

\section*{Conflict of interest} 
The authors declare that they have no conflict of interest.

\section*{Authors' contributions}
Equal contribution

% BibTeX users please use one of
\bibliographystyle{apalike}      % basic style, author-year citations
%\bibliographystyle{spmpsci}      % mathematics and physical sciences
%\bibliographystyle{spphys}       % APS-like style for physics
%\bibliographystyle{spbasic}
\bibliography{Bibliografia.bib}
\end{document}
% end of file template.tex

>>>>

\documentclass[review]{elsarticle}

\usepackage{lineno,hyperref}
\usepackage{amsfonts,amssymb,amsthm,amsmath,latexsym}  
\usepackage{natbib}
\usepackage{amssymb,amsmath,latexsym}
\modulolinenumbers[5]
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{obs}[theorem]{Observation}
\newtheorem{lem}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corolary}
\journal{Artificial Intelligence Journal}

%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%

%% Numbered
%\bibliographystyle{model1-num-names}

%% Numbered without titles
%\bibliographystyle{model1a-num-names}

%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}

%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}

%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}

%% APA style
\bibliographystyle{model5-names}\biboptions{authoryear}

%% AMA style
%\usepackage{numcompress}\bibliographystyle{model6-num-names}

%% `Elsevier LaTeX' style
\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\begin{frontmatter}

@@

%% Group authors per affiliation:
\author{Mauricio Gonzalez Soto}
\author{Luis E. Sucar Succar}
\author{Hugo J. Escalante Balderas}
\address{Instituto Nacional de Astrof\'isica \'Optica y Electr\'onica, Luis Enrique Erro 1, Santa Maria Tonanzintla, Mexico}

\begin{abstract}
Causal reasoning is a fundamental aspect of intelligent reasoning since it allows to consider interventions and counterfactuals. Decision making under uncertainty has been well studied when information is considered at the associative (probabilistic) level. The classical Theorems of von Neumann-Morgenstern and Savage provide a formal criterion for rational choice using purely associative information, which form the basis of many learning algorithms such as those used in Reinforcement Learning. In this work, we consider decision problems in which available actions and consequences are causally connected. We define a Causal Decision Problem and within this framework we state a previous result from J. Pearl, which relies on a known causal model and thus showing that it can be considered as a causal version of the classical von Neumann-Morgenstern Theorem. Furthermore, we consider the case when the causal mechanism that controls the environment is unknown to the decision maker, and propose and prove a causal version of Savage's Theorem. Then, we describe two applications for which these theorems provide theoretical foundations: causal games and optimal action learning in causal environments. These results highlight the importance of causal models in decision making and the variety of potential applications.
\end{abstract}

\begin{keyword}
Causality \sep Decision Making \sep Game Theory
\end{keyword}

\end{frontmatter}

\linenumbers

%\tableofcontents
\section{Introduction}
\label{intro}
Causal reasoning is a constant element in our lives as it is in human nature to constantly ask \textit{why} \citep{spirtes2000causation,pearl2018why}. Looking for causes is an everyday task and, in fact, causal reasoning is to be found at the very core of our minds \citep{tversky1977causal,waldmann2013causal,danks2014unifying,neil2019causality}. It has been argued that the brain is a causal inference machine which uses \textit{effects} to figure out \textit{causes} in order to engage with the world \citep{friston2010free,clark2015surfing}. 
Furthermore, acting in the world is conceived by human beings as \textit{intervening} the world and in fact humans are able to learn and use causal relations while making choices \citep{ tversky1980causal, lagnado2007beyond, hagmayer2008causal, hagmayer2009decision, hagmayer2017causality}.

@@

@@

@@

\subsection{Organization of the paper}
\label{organization}
This paper is organized as follows: first, in Section \ref{classical_theorems} we present the classical decision making theorems which rely on associative information in their formulation: Theorems \ref{vNM} and \ref{savage}; then, in Section \ref{causal_decision_problems} we present a brief introduction to Causation and Causal Graphical Models. In Section \ref{related_work} we present previous approaches both to Causal Decision Making as well as its applications to interactive optimal action learning: causal reinforcement learning. In Section \ref{causal_results} we state a von Neumann-Morgenstern Theorem for Causal Decision Making that is due to Pearl, which assumes a known causal structure. Our main result is Theorem \ref{causal_savage} which extends Pearl's result to the case of an unknown causal structure. Finally two applications of Theorem \ref{causal_savage} are described. First, a causal analogue of Nash Equilibrium for games in which actions and consequences are causally connected is defined. Then, in Section \ref{optimal_action} we describe some optimal-action learning methods for causal environments which are able to learn an \textit{optimal intervention} under different assumptions and conditions; in particular, we describe and provide an extension for one of such works. In Section \ref{limitations} we comment on the limitations of the assumptions we consider. Finally, in Section \ref{conclusion} we summarize the main contributions of our work.

@@

\section*{References}
\bibliographystyle{model5-names}\biboptions{authoryear}
\bibliography{mybibfile}

\end{document}
